Negative Selection in the 20th Century

### 2020-05-15 why-negative-selection.Rmd
Use 100 PCs from Abdel. Split out some data files in the drake plan.

```md
# TODO
* Improve imputation of parents' YOB.
* Selection bias. 
  - You could weight subjects by education level; sex; birth region?

@@

drake::loadd(famhist)
drake::loadd(rgs)
drake::loadd(score_names)
drake::loadd(reverse_code)

@@

pretty_names <- function (names) {
  reverse <- names %in% reverse_code
  names <- gsub("_", " ", names)
  names %<>% 
        strsplit(" ") %>% 
        map_chr(~paste0(
          toupper(substring(.x, 1, 1)), 
          substring(.x, 2),
          collapse = " "
        ))
  names[reverse] %<>% paste0("*")

@@

@@

We run regressions both with and without controls for the 40 top principal components
of the genetic data.

@@

To get a further insight into this we regress *siblings* and *children* on
individual principal components themselves. As Figure \@ref(fig:plot-pc-regs) shows,
there are many more large and significant effects  in siblings regressions.

```{r plot-pc-regs, fig.cap = "Effect of principal components on number of siblings/children. Absolute effect sizes are plotted. Each dot represents one bivariate regression. Points are jittered on the Y axis.", fig.height = 2}

drake::loadd(res_pcs)
n_regs <- as.double(nrow(res_pcs))

>>>>

```md
# TODO

* Use Abdel's 100 PCs, not UKBB's 40. DONE

* Look at geography, esp. in siblings regressions.

* Divide siblings regressions by own YOB, into early and late.
  - Use this to confirm the basic "selection decreasing" story

* Plot average scores over time, as an intro graph
  - Can we work out how to plot selection effects on the "same scale"?
  - Let X_i be i's polygenic score.
  - If E(X_parents) = 1/n \sum X_i in the parent's generation, then
    E(X_kids) = 1/n \sum X_i K_i, in the children's gernation, where K_i is 
    the number of kids i has. 
  - Up to a change of means, this is the same as the sample correlation between
    X and K. 
  - Do we calculate correlations per year, then plot? Probably too low N....
    Maybe calculate the correlation over the whole sample but then use
    it with each year's X_i profile

* How much does age at first live birth "account for" selection?

* Work more on weighting the data?

* control for age in n_partners regressions

* check for new 100 PCs

@@

drake::loadd(famhist)
drake::loadd(rgs)
drake::loadd(score_names)

@@

pretty_names <- function (names) {
  names <- gsub("_", " ", names)
  names %<>% 
        strsplit(" ") %>% 
        map_chr(~paste0(
          toupper(substring(.x, 1, 1)), 
          substring(.x, 2),
          collapse = " "
        ))

@@

```{r}
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

```{r plot-means-over-time, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name),) %>% 
      ggplot(aes(YOB, score, group = score_name)) + 
      geom_line() +
      facet_wrap(~score_name) +
      theme(strip.text = element_text(size = 7)) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")
```

@@

We run regressions both with and without controls for the 100 top principal
components of the genetic data.

@@

```{r}
drake::loadd(res_pcs)
n_regs <- as.double(nrow(res_pcs))
prop_sig <- res_pcs %>% 
      group_by(dep.var) %>% 
      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
      pull(sig, name = dep.var)
```

To get a further insight into this we regress *siblings* and *children* on
individual principal components. As Figure \@ref(fig:plot-pc-regs)
shows, effects are larger and more significant in siblings regressions. 
`r prop_sig[["N siblings"]]` principal components significantly predicted number of
siblings, while only `r prop_sig[["N children"]]` significantly predicted number
of children.

```{r plot-pc-regs, fig.cap = "Effect of principal components of genetic data on number of siblings/children. Absolute effect sizes are plotted. Each dot represents one bivariate regression. Points are jittered on the Y axis.", fig.height = 2}
